A fast algorithm for global code motion of congruent instructions

### 2018-02-23 cgo/paper.tex
sinkable contents

\subsection{Barriers}
\label{subsec:barriers}
Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. A basic-block in \LLVM{} IR is actually an
extended basic-block \cite{ebb} because there might be non-returning calls in
the middle of a basic-block. There might be other instructions like stores,
where compiler may not guarantee progress of control flow, all such instructions
are marked as barriers. A barrier will prevent any instruction occurring
lexically below it from getting hoisted. On the other hand, a sink barrier will
prevent any instruction above it from getting sunk. If there are no barriers in
a basic block, any instruction satisfying other legality checks can be sunk.

\begin{lstlisting}[caption={Compute hoist and sink barriers},captionpos=b]
// Compute both hoist and sink barriers
void computeBarriers()
  for each basic-block B in a function:
    barrier_found = false
    for each instruction I in B:
      if I does not guarantee progress:
        mark I as a barrier instruction
        barrier_found = true
        break;

@@

In the absence of context, as in our current implementation, some instructions
which might be safe are still classified as barriers, e.g., calls with missing
attributes.  Computing barriers allows for efficient removal of non-movable
candidates, this makes the legality check converge faster.

@@

\begin{lstlisting}[caption={Global Code Motion},captionpos=b]
void doGCM()
  Analyses available:
    Dominator Tree, DFS Numbering, Memory SSA
  Compute DJ-graph of function
  constructMergeSet(CFG)
  computeBarriers()
  do:
    Compute GVN of each expression in the function
    // Repeat hoisting if any candidates were hoisted.
  while (doCodeHoisting() > 0)
  // Code-sinking only once
  doCodeSinking()
\end{lstlisting}

\GCM{} basically consists of two parts, i.e., hoisting and sinking. This
implementation only moves congruent instructions which makes it easier to ensure
that we do not introduce a redundant computation along any path of execution. If
the dependency of a hoistable candidate is in the same basic-block as the
candidate, then the dependency must also be hoistable, otherwise hoisting will
be illegal or would require a complicated code generation to make it legal. The
current algorithm discards cases if the dependency is neither hoistable nor
rematerializable. We collect the \GVN{} of all the instructions in the function
and iterate on the list of instructions having identical \GVN{}s. The algorithm
\emph{doGCM} prefers hoisting (\emph{doCodeHoisting}) to sinking
(\emph{doCodeSinking}).

\begin{lstlisting}[caption={Hoist instructions},captionpos=b]
int doCodeHoisting(VNs)
  // VNs = map (VN -> list of I with value VN)
  Sort VNs in ascending order

@@

\begin{lstlisting}[caption={Sink instructions},captionpos=b]
int doCodeSinking()
  // VNs = map (VN -> list of I with value VN)
  Sort VNs in ascending order

@@

After no more candidates are hoistable, sinking is performed. It is only
performed once because it may not open new opportunities for sinkable
candidates. For sinking we iterate on the $\Phi$ nodes. After each code-motion
the \SSA{} form is restored by updating the intermediate representation (IR) to
reflect the changes. At the same time \MemorySSA{} is updated as well. Both the
data structures can be updated very efficiently. After one full iteration of
hoisting, the \GVN{} needs to be recomputed to explore new hoistable candidates.
Finally after the transformation is done, we verify a set of post-conditions to
establish that program invariants are maintained: e.g., consistency of use-defs,
and \SSA{} semantics.

>>>>

\subsection{Barriers}
\label{subsec:barriers}
Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. A basic-block in \LLVM{} IR is actually an
extended basic-block \cite{ebb} because there might be non-returning calls in
the middle of a basic-block. There might be other instructions like stores,
where compiler may not guarantee progress of control flow, all such instructions
are marked as barriers as illustrated in Algorithm~\ref{algo:barriers}.

\begin{lstlisting}[caption={Compute hoist and sink barriers},captionpos=b,label={algo:barriers}]
// Compute both hoist and sink barriers
void computeBarriers()
  for each basic-block B in a function:
    barrier_found = false
    for each instruction I in B:
      if I does not guarantee progress:
        mark I as a barrier instruction
        barrier_found = true
        break;

@@

A barrier will prevent any instruction occurring lexically below it from getting
hoisted. On the other hand, a sink barrier will prevent any instruction above it
from getting sunk. If there are no barriers in a basic block, any instruction
satisfying other legality checks can be sunk.  In the absence of interprocedural
context, as in our current implementation, some instructions which might be safe
are still classified as barriers, e.g., calls with missing attributes.
Computing barriers allows for efficient removal of non-movable candidates, this
makes the legality check converge faster.

@@

\begin{lstlisting}[caption={Global Code Motion},captionpos=b,label={algo:gcm}]
void doGCM()
  Analyses available: Dominator Tree, DFS Numbers, Memory SSA
  Compute DJ-graph of CFG
  constructMergeSet(CFG)
  computeBarriers()
  do:
    Compute GVN of each expression in the function
    // Repeat hoisting if any candidates were hoisted.
  while (doCodeHoisting() > 0)
  doCodeSinking() // Code-sinking only once
\end{lstlisting}

\GCM{} (Algorithm~\ref{algo:gcm}) basically consists of two parts, i.e., hoisting
(Algorithm~\ref{algo:hoist}) and sinking (Algorithm~\ref{algo:sink}). This
implementation only moves congruent instructions which makes it easier to ensure
that we do not introduce a redundant computation along any path of execution. If
the dependency of a hoistable candidate is in the same basic-block as the
candidate, then the dependency must also be hoistable, otherwise hoisting will
be illegal or would require a complicated code generation to make it legal. The
current algorithm discards cases if the dependency is neither hoistable nor
rematerializable. We collect the \GVN{} of all the instructions in the function
and iterate on the list of instructions having identical \GVN{}s. The algorithm
\emph{doGCM} prefers hoisting (\emph{doCodeHoisting}) to sinking
(\emph{doCodeSinking}).

\begin{lstlisting}[caption={Hoist instructions},captionpos=b,label={algo:hoist}]
int doCodeHoisting(VNs)
  // VNs = map (VN -> list of I with value VN)
  Sort VNs in ascending order

@@

After no more candidates are hoistable, sinking is performed. To sink the
instructions $\Phi$ (Section~\ref{fig:phi-intro}) nodes are considered as they
keep track of expressions which are available at the point of their definition.
Sinking is only performed once because there were far fewer sinkable
instructions than hoistable instructions during our experiments.

Moreover, as the direction of movement of sinkable instruction is in the same
direction as program flow, the sinkable instructions which are dependent on
other sinkable instructions can be sunk in the same iteration as algorithm
iterates bottom up.

\begin{lstlisting}[caption={Sink instructions},captionpos=b,label={algo:sink}]
int doCodeSinking()
  // VNs = map (VN -> list of I with value VN)
  Sort VNs in ascending order

@@

After each code-motion the \SSA{} form is restored by updating the
intermediate representation (IR) to reflect the changes. At the same time
\MemorySSA{} is updated as well. Both the data structures can be updated very
efficiently. After one full iteration of hoisting, the \GVN{} needs to be
recomputed to explore new hoistable candidates.  Finally after the
transformation is done, we verify a set of post-conditions to establish that
program invariants are maintained: e.g., consistency of use-defs, and \SSA{}
semantics.
